---
title: A Psychphysics Study on Glance reading accuracy in the peripheral field
author:
  - name: Martin Schmettow
    id: ut
    orcid: 0000-0003-0240-4087
    email: m.schmettow@utwente.nl
    affiliation: 
      - name: University of Twente
        city: Enschede, NL
        url: www.utwente.nl
    attributes:
        corresponding: true
keywords:
  - Psychophysics
format: docx
bibliography: references.bib
date: last-modified
execute:
  echo: false
  warning: false
  message: false
  cache: true
---
  
# Data reading

```{r}
#! eval: false
library(tidyverse)

csv_files <- dir("data", pattern = "*.csv", full.names = T)

read_gr <- function(f) {
  read_csv(f, show_col_types = FALSE) |> 
    filter(!is.na(xPos)) |> 
    select(participant, xPos, stimulusSet, whichImage, thisN, blockNumber, direction, angle, stimulus, response, correct) |> 
#    mutate(blockNumber = as.factor(blockNumber)) |> 
    filter(!is.na(blockNumber))
}

T_blocks <-
  tibble(blockNumber = c(1,2,3),
         eccentricity = c(5, 10, 15))


D_0 <- 
  csv_files |> 
  map_df(read_gr)

read_gr(csv_files[1])

D_1 <- 
  D_0 |> 
  left_join(T_blocks, by = join_by(blockNumber)) |> 
  mutate(Part = as.factor(participant)) |>
  group_by(Part) |> 
  mutate(trial = row_number()) |>
  ungroup() |> 
  select(Part, trial, angle, eccentricity, correct) |> 
  bayr::as_tbl_obs()

summary(D_1)

save(D_0, D_1, file = "Data.Rda")
```

# Exploration

The following plot shows the population level proportion of correct responses as a function of the angle and eccentricity. There is huge fluctuations with angles below 10 degree, which is due to the adaptive staircase procedure, which results in fewer observations for these angles. The tendency above 10 degree is 

```{r}
#| fig-width: 8
#| fig-height: 12

T_1 <- D_1 |> 
  summarize(p_correct = mean(correct),
            n_Obs = n(),
            .by = c(angle, eccentricity))
  
T_1 |> 
  ggplot(aes(x = angle, color = as.factor(eccentricity), y = p_correct)) +
  geom_point(aes(size = n_Obs )) +
  geom_smooth(se = F, method = "loess")


  
```

The figure below shows the proportion of correct responses as a function of the Angle (in three ranges) and eccentricity, separately for each participant.

```{r}
#| fig-width: 8
#| fig-height: 12

D_1 |> 
  mutate(Angle = case_when(
    angle < 10 ~ "0-10",
    angle < 15 ~ "10-15",
    angle <= 20  ~ "15-20"
  )) |>
  summarize(p_correct = mean(correct),
            n_Obs = n(),
            .by = c(Part, Angle, eccentricity)) |> 
  ggplot(aes(x = Angle, color = as.factor(eccentricity), y = p_correct)) +
  facet_wrap(~Part) +
  geom_point(aes(size = n_Obs), alpha = 0.5) +
  geom_line(aes(group = eccentricity), alpha = 0.5)
  
```

## Conclusion

Many participants even the easiest condition don't improve beyond chance level. Only in a few do we observe a regular pattern that matches our expectations. Still, a subset of participants can perhaps be used with the model. Legit participants should show performance above chance level and have clear upwards trend, as well as the expected ordering of eccentricity.


# Modelling

The theoretically complete Psychophysics function would have:
  + dprime
  + a guessing rate
  + a lapse rate (readout error)
  + a decision bias
  
The theoretical base for lapse is that the response depends on the sensory characteristics and the readout process  [@Gold2013].


```{r}
inv_logit <- function(x) {
  exp(x) / (1 + exp(x))
}

logit <- function(x) {
  log(x / (1 - x))
}

pm_func <- function(x, dprime, lapse, bias, guess = 0.5) {
  guess + (1 - guess - lapse) * inv_logit(dprime * x - bias)
}
```

```{r}
tibble(
  x = seq(-10, 10, length.out = 100),
  y = pm_func(x, dprime = 0.5, lapse = 0.05, bias = 0.6)
) |> 
  ggplot(aes(x, y)) +
  geom_line()
```



```{r}
#| eval: false
# devtools::install_github("schmettow/bayr")

```

The lapse rate and decision bias are modeled as functions of the participant, but not of the eccentricity, because it is assumed to be a readout error that is independent of the stimulus. Dprime is modelled as a function of eccentricity.

Since we have no data at the upper end of performance, the lapse parameter cannot be estimated neither on  participant level nor on population level. The same holds for the bias parameter. Therefore, a much simplified model is used.

```{r}
#| eval: false
library(tidyverse)
library(brms)

F_1 <- bf(
  correct ~ 0.5 + (1 - 0.5) * inv_logit(exp(dprime) * angle), 
  dprime ~ 1 + eccentricity + (1 + eccentricity | Part),
  #lapse ~ 1, 
  nl = TRUE
)

Priors_1 <- c(
  set_prior("normal( 0, 1)", nlpar = "dprime")
  #set_prior("normal( 0, 0.5)", nlpar = "bias"),
  #set_prior("normal(-2.2, 0.6)", nlpar = "lapse")
)

M_1 <- brm(
  F_1,
  data = D_1,
  family = bernoulli(link = "identity"),
  prior = Priors_1,
  chains = 8,
  cores = 8,
  iter = 3000,
  warmup = 2000,
  control = list(adapt_delta = 0.99999999, max_treedepth = 12)
)

save(M_1, D_1, file = "M_1.Rda")

```

Here we extract the posterior samples. 

```{r}
library(bayr)
load("M_1.Rda")
P_1 <- bayr::posterior(M_1)

```

This extracts the population-level parameters. Parameters are transformed back to their original scale. Note that by the back-transformation `dprime` is a *multiplicative* parameters.

```{r}
fixef(P_1)
```


This figure shows the posterior distributrions of population-level parameters.

```{r}
P_1 |> 
  filter(type == "fixef") |> 
  ggplot(aes(x = value, col = parameter)) +
  facet_grid(~nonlin) +
  geom_density()

```


## Posterior predictions

```{r}
PP_1 <- bayr::post_pred(M_1, ndraws = 1000)
```

